Notes on filtration of logics containing {\bf K5}

### 2021-04-20 filtration-stuff/NewText.tex
GL and finite canonical models

The converse implication does not have to true generally. We prove a bit more weak version of this implication. A logic admits $\mathcal{L}$ admits $\Diamond$-filtrations, if for each model $\mathcal{M} \models \mathcal{L}$ and for each $\Gamma$, a $\operatorname{Sub}$-closed set of formulas, there exists a model $\widehat{\mathcal{M}}$, a filtration of $\mathcal{L}$ through $\operatorname{Sub}(\Gamma \cup \Diamond \Gamma)$.

${\bf K}5$ is an example of such a logic since ${\bf K}5$ admits filtrations through sets having the form $\Gamma \cup \{ \Diamond \Box \psi \: | \: \Box \psi \in \Gamma \}$.

Note that, if $\mathcal{L}$ admits $\Diamond$-filtrations, so $\mathcal{L}$ admits the ``finite canonical'' model property, the proof is the same as in Lemma~\ref{good}.

\begin{lemma}
  Let $\mathcal{L}$ be a logic that admits $\Diamond$-filtrations and $\Gamma$ a $\operatorname{Sub}$-closed set of formulas, then there exists a ``finite canonical'' model $\mathcal{M}^{\Gamma}_{\mathcal{L}} \models \mathcal{L}$
\end{lemma}

\begin{proof}
Let $\Gamma$ be a finite $\operatorname{Sub}$-closed of formulas and $\mathcal{M}_{\mathcal{L}}^{\Gamma} = \langle W_{\mathcal{L}}^{\Gamma}, R_{\mathcal{L}}^{\Gamma}, \vartheta_{\mathcal{L}}^{\Gamma} \rangle$ a ``finite canonical'' model generated by $\Gamma$.

Let us show that there exists a model $\widehat{\mathcal{M}}$, a strict filtration of $\mathcal{M}$ through $\operatorname{Sub}(\Gamma \cup \Diamond \Gamma) := \Psi$.

Let $\sim$ be an equivalence relation on $\mathcal{M}$ defined as usual:

    \begin{center}
      $u \sim v$ iff $\forall \psi \in \Psi$
      $\mathcal{M}, u \models \psi \Leftrightarrow \mathcal{M}, v \models \psi$
    \end{center}

    Let $[u] \in \mathcal{M}/\sim$, consider its truth set $||[u]|| = \{ \psi \in \Psi \: | \: \mathcal{M}, u \models \psi \}$.
    $||[u]||$ is clearly a finite $\mathcal{L}$-consistent set and it is maximal.
    Consider a map:
    \begin{center}
    $\iota : W/\sim \to W_{\mathcal{L}}^{\Gamma}$ with $\iota : [u] \mapsto ||[u]||$
    \end{center}

    Consider the set $\iota(W/\sim)$ that we denote as $\operatorname{Im}(\iota)$. Let us show that $\operatorname{Im}(\iota)$ is $R_{\mathcal{L}}^{\Gamma}$-closed.

    Let $\iota([u]) R_{\mathcal{L}}^{\Gamma} \Delta$, let us show that $\Delta \in \iota(W/\sim)$, that is, $\Delta = \iota([v])$ for some $v \in W$.

    Note that $\bigwedge \limits_{\psi \in \Delta} \psi \in \Delta$ since $\Delta$ is finite.
    This conjuction has the form $\bigwedge \limits_{\psi \in \Delta} \psi \in \Delta = \psi_1 \land \dots \land \psi_n$ for some $n < \omega$ and for some $\psi_i \in \Gamma \cup \neg \Gamma$, where $0 < i < \omega$.

    Therefore $\mathcal{M}^{\Gamma}_{\mathcal{L}}, \iota([u]) \models \Diamond (\psi_1 \land \dots \land \psi_n)$.
    Then $\mathcal{M}, u \models \Diamond (\psi_1 \land \dots \land \psi_n)$. That implies $\mathcal{M}, v \models \psi_1 \land \dots \land \psi_n$ for some $v \in R(u)$.
    This is clearly a desired $v$ since $||[v]|| \subseteq \Delta$ and such a $\Delta$ is unique.

    Thus, $\langle \operatorname{Im}(\iota), R_{\mathcal{L}}^{\Gamma}, {\vartheta'}_{\mathcal{L}}^{\Gamma} \rangle$, where ${\vartheta'}_{\mathcal{L}}^{\Gamma}(p) = \vartheta_{\mathcal{L}}^{\Gamma}(p) \cap \operatorname{Im}(\iota)$
    is a generated submodel of $\mathcal{M}_{\mathcal{L}}^{\Gamma}$.

    Consider a model $\widehat{\mathcal{M}} = \langle W/\sim, \widehat{R}, \widehat{\vartheta} \rangle$, where $\widehat{R}$ is defined as
    \begin{center}
      $[u] \widehat{R} [v]$ iff $||[u]|| {R}_{\mathcal{L}}^{\Gamma} ||[v]||$
    \end{center}

    Let us show that $R_{min} \subseteq \widehat{R} \subseteq R_{max}$.

    Suppose that there are $u' \in [u]$ and $v' \in [v]$ with $u' R v'$. Let us show that
    $[u] \widehat{R} [v]$. That is, let us show that $\Diamond \iota(v) \subseteq \iota(u)$.

    Suppose $\psi \in \iota(v)$, that is, $\psi \in ||v||$, that denotes $\psi \in \Psi$ and $\mathcal{M}, v \models \psi$. Therefore, $\mathcal{M}, v' \models \psi$. Then $\mathcal{M}, u' \models \Diamond \psi$. $\Diamond \psi \in \Psi$ and $u \sim_{\Gamma} u'$, so
    $\mathcal{M}, u \models \Diamond \psi$.

    The second inclusion follows from the definition of a relation in a finite ``canonical'' model.
\end{proof}

In particular, ${\bf K}5$ has the finite ``canonical'' model property.

>>>>

The converse implication does not have to true generally.

Recall that ${\bf GL} = {\bf K} \oplus \Diamond (p \land \neg \Diamond p) \to \Diamond p$. ${\bf GL}$ does not admit filtrations, but one may prove that ${\bf GL}$ has the FMP using the selective filtation technique.

\begin{defin}
  Let $\mathcal{M} = \langle W, R, \vartheta \rangle$ be a model and $\Gamma$ a set closed under subformulas, then a weak submodel $\mathcal{M} = \langle W', R', \eta \rangle$ (that is, $W' \subseteq W$, $R' \subseteq R$, $\eta = \vartheta \upharpoonright_{W'}$) is called a selective filtration of $\mathcal{M}$ through $\Gamma$ is the following holds, for $\Diamond \psi \in \Gamma$, $x \in W'$:
  \begin{center}
    $\mathcal{M}, x \models \Diamond \psi$ iff $\mathcal{M}, y \models \psi$ for some $y \in R'(x)$
  \end{center}
\end{defin}

\begin{theorem}
  ${\bf GL}$ admits the ``finite canonical'' model property.
\end{theorem}

\begin{proof}
  Let $\Gamma$ be a finite $\operatorname{\bf Sub}$-closed set. So, $\Gamma = \operatorname{\bf Sub}(\gamma_1) \cup \dots \operatorname{\bf Sub}(\gamma_n)$, for some $\gamma_1, \dots, \gamma_n \in \Gamma$ and every $\gamma_i$ is ${\bf GL}$-consistent.

  $\gamma$ is true in $\mathcal{M}_{\Gamma} = \langle V, R^{\bf GL}, \vartheta \rangle$, where
  \begin{center}
  $V_{\Gamma} = \bigcup \limits_{\psi \in \Gamma} \{ x \in W_{\bf GL} \: | \: \mathcal{M}_{\bf GL}, x \models \psi \land \neg \Diamond \psi \}$
  \end{center}
  since $\mathcal{M}_{\Gamma}$ is a selective filtration of $\mathcal{M}_{\bf M}$ through $\Gamma$.

  Let $x \in V$, consider $||x|| = \{ \psi \in \Gamma \: | \: \mathcal{M}_{\bf GL}, x \models \psi \}$
  and define $x R^{\bf GL} y$ implies $||x|| R_{\Gamma} ||y||$. Every $||x||$ is clearly finite.
  The valuation map is usual, $\vartheta_{\Gamma}(p) = \{ ||x|| \: | \: p \in ||x|| \}$ for $p \in \operatorname{PV} \cap \Gamma$. We put $W_{\Gamma}$

  Let us check that $\langle W_{\Gamma}, R_{\Gamma}, \vartheta_{\Gamma} \rangle$ is a ``finite canonical`` ${\bf GL}$-model.
  Let us check that $R_{\Gamma}$ is transitive and Noetherian and it has no relflexive points.
  If $W_{\Gamma}$ had a relfexive point, say $||y||$, then we would have $\mathcal{M}, y \models \psi \to \Diamond \psi$, but that is not impossible. Transitivity is obvious.

  Clearly that $x R^{\bf GL} y$ implies $||x|| R_{\Gamma} ||y||$ and vice versa. So, if we had an infinite decreasing $R_{\Gamma}$-chain, then we would have an infinite decreasing $R^{\bf GL}$-chain, but that is not possible.
  So $\langle W_{\Gamma}, R_{\Gamma}, \vartheta_{\Gamma} \rangle$ is a ``finite canonical`` ${\bf GL}$-model.
\end{proof}

So we have an example of a logic that has the ``finite canonical'' model property with no filtrations.
